Impact of Training Variability on Visuomotor Function Learning and Extrapolation

### 2024-06-22 Analysis/e1_Analysis.qmd
minor changes

::: {.column-page-right}
```{r}

@@

tables <- map(vars_to_plot, function(pvar) {
  testAvg |>
    group_by(condit, vb) |>
    summarize(mean = mean(.data[[pvar]]), sd = sd(.data[[pvar]])) |>
    mutate(Distribution = vb) |>
    gt() |>
    tab_spanner(
      label = 'Training Condit',
      columns = -condit
    ) |>
    cols_label_with(fn = str_to_title) |>
    fmt_number(decimals = 2) |>
    cols_align('left', columns = condit) |>
    text_transform(
      locations = cells_body(columns = 'Distribution'),
      fn = function(column) {
        map2(unique_combinations$condit, unique_combinations$vb, ~plot_violin_condit(.x, .y, pvar)) |>
          ggplot_image(height = px(50), aspect_ratio = 3)
      }
    )
})
#print(tables[[1]])
#print(tables[[2]])

@@

```{r}
#| fig-cap: "Mean Vx over training blocks"
#| fig-width: 10
#| fig-height: 9
#| tbl-cap: "test table"
#| eval: false
# vpt1=plotWithTable(vp1,vt1,arrange="V")
nb=5
vp1=e1 |> filter(expMode=="train") |> learn_curve_plot(gt.train,vx,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb)
vp1

@@

```{r}
#| fig-cap: "Absolute Deviation from Band over training"
#| fig-width: 8
#| eval: false
#| 
vp2=e1 |> filter(expMode=="train") |> 
  learn_curve_plot(gt.train,dist,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb) 
vp2

@@

```{r}
#| fig-cap: "Signed Deviation from Band over training"
#| fig-width: 8
#| eval: false
vp3=e1 |> filter(expMode=="train") |> learn_curve_plot(gt.train,sdist,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb) 
vp3

@@

# 22
::: {.column-page-right}
```{r}
#| layout: [[65, 65],[-5], [1,550, 550]]
#| label: train tables2
#| echo: fenced
#| tbl-subcap: 
#|   - "Vx"
#|   - "Deviation from target"
#| eval: false
#| 
vt1 %>% kable(format = "html",escape = FALSE)
vt2 %>% kable(format = "html",escape = FALSE)

@@

# 33
```{r}
#| layout: [[65, 65],[-5], [1,350, 350,10]]
#| label: train tables
#| echo: fenced
#| tbl-subcap: 
#|   - "Vx"
#|   - "Deviation from target"
#| column: screen-inset-right
#| eval: false
vt1 %>% kable(format = "html",escape = FALSE)
vt2 %>% kable(format = "html",escape = FALSE)

@@

```{r}
#| label: tbl-example4
#| tbl-cap: "Example"
#| tbl-subcap: 
#|   - "Signed Distance"
#|   - "Percent Hit"
#| layout: [[80, 90],[2], [-60,650]]
#| echo: fenced
#| column: screen-inset-right
#| fig-width: 14
#| eval: false

@@

```{r}
#| fig-width: 12
#| fig-height: 11
#| column: page-inset-right
#| eval: false
((vp1+big_text()+vp2+big_text())/(vp3+big_text()+vp4+big_text())) + plot_layout(guides = "collect")

>>>>

@@

tables <- map(vars_to_plot, function(pvar) {
  testAvg |>
    group_by(condit, vb) |>
    summarize(mean = mean(.data[[pvar]]), sd = sd(.data[[pvar]])) |>
    mutate(Distribution = vb) |>
    gt() |>
    tab_spanner(
      label = 'Training Condit',
      columns = -condit
    ) |>
    cols_label_with(fn = str_to_title) |>
    fmt_number(decimals = 2) |>
    cols_align('left', columns = condit) |>
    text_transform(
      locations = cells_body(columns = 'Distribution'),
      fn = function(column) {
        map2(unique_combinations$condit, unique_combinations$vb, ~plot_violin_condit(.x, .y, pvar)) |>
          ggplot_image(height = px(50), aspect_ratio = 3)
      }
    )
})
print(tables[[1]])
print(tables[[2]])

@@

```{r}
#| fig-cap: "Mean Vx over training blocks"
#| fig-width: 10
#| fig-height: 9
#| tbl-cap: "test table"

# vpt1=plotWithTable(vp1,vt1,arrange="V")
nb=5
vp1=e1 |> filter(expMode=="train") |> learn_curve_plot(gt.train,vx,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb)
vp1

@@

```{r}
#| fig-cap: "Absolute Deviation from Band over training"
#| fig-width: 8

vp2=e1 |> filter(expMode=="train") |> 
  learn_curve_plot(gt.train,dist,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb) 
vp2

@@

```{r}
#| fig-cap: "Signed Deviation from Band over training"
#| fig-width: 8

vp3=e1 |> filter(expMode=="train") |> learn_curve_plot(gt.train,sdist,condit,facet_var=vb,groupVec=c(gt.train,condit,tOrder,id,vb),nbins=nb) 
vp3

@@

# 22
::: {.column-page-right}
```{r}
#| layout: [[65, 65],[-5], [1,550, 550]]
#| label: train tables2
#| echo: fenced
#| tbl-subcap: 
#|   - "Vx"
#|   - "Deviation from target"

vt1 %>% kable(format = "html",escape = FALSE)
vt2 %>% kable(format = "html",escape = FALSE)

@@

# 33
```{r}
#| layout: [[65, 65],[-5], [1,350, 350,10]]
#| label: train tables
#| echo: fenced
#| tbl-subcap: 
#|   - "Vx"
#|   - "Deviation from target"
#| column: screen-inset-right

vt1 %>% kable(format = "html",escape = FALSE)
vt2 %>% kable(format = "html",escape = FALSE)

@@

```{r}
#| label: tbl-example4
#| tbl-cap: "Example"
#| tbl-subcap: 
#|   - "Signed Distance"
#|   - "Percent Hit"
#| layout: [[80, 90],[2], [-60,650]]
#| echo: fenced
#| column: screen-inset-right
#| fig-width: 14
#|

@@

```{r}
#| fig-width: 12
#| fig-height: 11
#| column: page-inset-right
((vp1+big_text()+vp2+big_text())/(vp3+big_text()+vp4+big_text())) + plot_layout(guides = "collect")

### 2024-06-22 Manuscript/manuscript.qmd
minor changes

### Variability and Function Learning

@@

  data1 [label = " Varied Training \n800-1000\n1000-1200\n1200-1400", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n800-1000", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 7.0
//| fig-height: 2.2
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#FF0000"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#00A08A"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```

>>>>

## Variability and Function Learning

@@

  data1 [label = " Varied Training \n800-1000\n1000-1200\n1200-1400", fillcolor = "#00A08A"]
  data2 [label = " Constant Training \n800-1000", fillcolor = "#FF0000"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]

@@

::: {.content-visible unless-format="markdown"}
```{dot}
//| label: fig-design-e2
//| fig-cap: "Experiment 2 Design. Constant and Varied participants complete different training conditions. The training and testing bands are the reverse of Experiment 1. "
//| fig-width: 7.0
//| fig-height: 2.2
//| fig-responsive: false
//| echo: false 
//| 
/////| column: screen-inset-right
digraph {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = filled]
  data1 [label = " Varied Training \n100-300\n350-550\n600-800", fillcolor = "#00A08A"]
  data2 [label = " Constant Training \n600-800", fillcolor = "#FF0000"]
  Test3 [label = "    Final Test \n  Novel With Feedback  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
  data1 -> Test1
  data2 -> Test1
  subgraph cluster {
    label = "Test Phase \n(Counterbalanced Order)"
    Test1 [label = "Test  \nNovel Bands  \n800-1000\n1000-1200\n1200-1400", fillcolor = "#ECCBAE"]
    Test2 [label = "  Test \n  Varied Training Bands  \n100-300\n350-550\n600-800", fillcolor = "#ECCBAE"]
    Test1 -> Test2
  }
  Test2 -> Test3
}
```

### 2024-06-22 Sections/htw_full.qmd
minor changes

---
title: Impact of Training Variability on Visuomotor Function Learning and Extrapolation
author:
- name: Thomas E. Gorman
  affiliations: 
   - Department of Psychological and Brain Sciences, Indiana University-Bloomington
  affiliation-url: https://psych.indiana.edu/
  url: https://tegorman13.github.io/
  email: tegorman@iu.edu
  orcid: 0000-0001-5366-5442
- name: Robert L. Goldstone
  affiliation: Department of Psychological and Brain Sciences, Indiana University-Bloomington
  affiliation-url: https://psych.indiana.edu/
  url: https://pc.cogs.indiana.edu/
  email: rgoldsto@indiana.edu
  orcid: 0000-0001-8357-8358
page-layout: full
code-fold: true
code-tools: true
cache: true
toc: true
toc-location: body
toc-title: Contents
repo-actions: false
execute: 
  warning: false
  eval: true
format:
  html: 
    date: last-modified
  hugo-md:
    echo: false
    include: true
    html-math-method: mathjax
    output-file: htw_hugo.md
  gfm:
    echo: false
    #output: false
    output-file: htw_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "htw.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
  docx:
    echo: false
    prefer-html: true
    output-file: "htw.docx"
    toc: true
prefer-html: true
format-links: [hikmah-manuscript-pdf, gfm, docx]
---

@@

The study of human function learning investigates how people learn relationships between continuous input and output values.  Function learning is studied both in tasks where individuals are exposed to a sequence of input/output pairs [@deloshExtrapolationSineQua1997; @mcdanielEffectsSpacedMassed2013], or situations where observers are presented with an incomplete scatterplot or line graph and make predictions about regions of the plot that don't contain data [@ciccioneCanHumansPerform2021; @courrieuQuickApproximationBivariate2012; @saidExtrapolationAccuracyUnderestimates2021;@schulzCommunicatingCompositionalPatterns2020].

@carrollFunctionalLearningLearning1963 conducted the earliest work on function learning. Input stimuli and output responses were both lines of varying length. The correct output response was related to the length of the input line by a linear, quadratic, or random function.  Participants in the linear and quadratic performed above chance levels during extrapolation testing, with those in the linear condition performing the best overall. Carroll argued that these results were best explained by a ruled based model wherein learners form an abstract representation of the underlying function. Subsequent work by @brehmerHypothesesRelationsScaled1974,testing a wider array of functional forms, provided further evidence for superior extrapolation in tasks with linear functions. Brehmer argued that individuals start out with an assumption of a linear function, but  given sufficient error will progressively test alternative hypothesis with polynomials of greater degree. @kohFunctionLearningInduction1991 employed a visuomotor function learning task, wherein participants were trained on examples from an unknown function relating the length of an input line to the duration of a response (time between keystrokes). In this domain, participants performed best when the relation between line length and response duration was determined by a power, as opposed to linear function. Koh & Meyer developed the log-polynomial adaptive-regression model to account for their results.

@@

*Model Fits to Experiment 2 and 3.* Data from Experiments 2 and 3 were fit to ALM and EXAM in the same manner as Experiment1 . For brevity, we only plot and discuss the results of the "fit to training and testing data" models - results from the other fitting methods can be found in the appendix. The model fitting results for Experiments 2 and 3 closely mirrored those observed in Experiment 1. The Bayesian regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and their interaction (see @tbl-htw-ee-e23) revealed a consistent main effect of Model across all three experiments. The negative coefficients for the ModelEXAM term (Exp 2: $\beta$ = `r bm1_e2$estimate`, 95% CrI `r bm1_e2$conf.low`, `r bm1_e2$conf.high`, pd = `r bm1_e2$pd`; Exp 3: $\beta$ = `r bm1_e3$estimate`, 95% CrI `r bm1_e3$conf.low`, `r bm1_e3$conf.high`, pd = `r bm1_e3$pd`) indicate that EXAM outperformed ALM in both experiments. Furthermore, the interaction between Model and Condition was significant in both Experiment 2 ($\beta$ = `r bm3_e2$estimate`, 95% CrI `r bm3_e2$conf.low`, `r bm3_e2$conf.high`, pd = `r bm3_e2$pd`) and Experiment 3 ($\beta$ = `r bm3_e3$estimate`, 95% CrI `r bm3_e3$conf.low`, `r bm3_e3$conf.high`, pd = `r bm3_e3$pd`), suggesting that the superiority of EXAM over ALM was more pronounced for the Constant group compared to the Varied group, as was the case in Experiment 1. Recall that Experiment 3 included participants in both the original and reverse order conditions - and that this manipulation interacted with the effect of training condition. We thus also controlled for band order in our Bayesian Regression assessing the relative performance of EXAM and ALM in Experiment 3. There was a significant three way interaction between Model, Training Condition, and Band Order ($\beta$ = `r bm4_e3$estimate`, 95% CrI `r bm4_e3$conf.low`, `r bm4_e3$conf.high`, pd = `r bm4_e3$pd`), indicating that the relative advantage of EXAM over ALM was only more pronounced in the original order condition, and not the reverse order condition (see @fig-e2_e3_ae).

@@

*Computational Model Summary*. Across all three experiments, the model fits consistently favored the Extrapolation-Association Model (EXAM) over the Associative Learning Model (ALM). This preference for EXAM was particularly pronounced for participants in the constant training conditions (note the positive coefficients on ModelEXAM:conditVaried interaction terms @tbl-htw-ee-e23). This pattern is clearly illustrated in @fig-htw-best-model, which plots the difference in model errors between ALM and EXAM for each individual participant. Both varied and constant conditions have a greater proportion of subjects better fit by EXAM (positive error differences), with the magnitude of EXAM's advantage visibly larger for the constant group.

>>>>

---
title: Impact of Training Variability on Visuomotor Function Learning and Extrapolation
author:
- name: Thomas E. Gorman
  affiliations: 
  - name:  Department of Psychological and Brain Sciences, Indiana University-Bloomington, USA
    affiliation-url: https://psych.indiana.edu/
  - name:  Program in Cognitive Science, Indiana University-Bloomington, USA
    affiliation-url: https://cogs.indiana.edu/index.html
  url: https://tegorman13.github.io/
  email: tegorman@iu.edu
  orcid: "0000-0001-5366-5442"
- name: Robert L. Goldstone
  affiliation: Department of Psychological and Brain Sciences, Indiana University-Bloomington
  affiliation-url: https://psych.indiana.edu/
  url: https://pc.cogs.indiana.edu/
  email: rgoldsto@indiana.edu
  orcid: 0000-0001-8357-8358
page-layout: full
code-fold: true
code-tools: true
cache: true
toc: true
toc-location: body
toc-title: Contents
repo-actions: false
execute: 
  warning: false
  eval: true
format:
  html: 
    date: last-modified
  hugo-md:
    echo: false
    include: true
    html-math-method: mathjax
    output-file: htw_hugo.md
  gfm:
    echo: false
    #output: false
    output-file: htw_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "htw.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
  docx:
    echo: false
    prefer-html: true
    output-file: "htw.docx"
    toc: true
prefer-html: true
format-links: [hikmah-manuscript-pdf, gfm, docx]
---

@@

The study of human function learning investigates how people learn relationships between continuous input and output values.  Function learning is studied both in tasks where individuals are exposed to a sequence of input/output pairs [@deloshExtrapolationSineQua1997; @mcdanielEffectsSpacedMassed2013], or situations where observers are presented with an incomplete scatterplot or line graph and make predictions about regions of the plot that do not contain data [@ciccioneCanHumansPerform2021; @courrieuQuickApproximationBivariate2012; @saidExtrapolationAccuracyUnderestimates2021;@schulzCommunicatingCompositionalPatterns2020].

@carrollFunctionalLearningLearning1963 conducted the earliest work on function learning. Input stimuli and output responses were both lines of varying length. The correct output response was related to the length of the input line by a linear, quadratic, or random function.  Participants in the linear and quadratic performed above chance levels during extrapolation testing, with those in the linear condition performing the best overall. Carroll argued that these results were best explained by a rule-based model wherein learners form an abstract representation of the underlying function. Subsequent work by @brehmerHypothesesRelationsScaled1974,testing a wider array of functional forms, provided further evidence for superior extrapolation in tasks with linear functions. Brehmer argued that individuals start out with an assumption of a linear function, but  given sufficient error will progressively test alternative hypothesis with polynomials of greater degree. @kohFunctionLearningInduction1991 employed a visuomotor function learning task, wherein participants were trained on examples from an unknown function relating the length of an input line to the duration of a response (time between keystrokes). In this domain, participants performed best when the relation between line length and response duration was determined by a power, as opposed to linear function. Koh & Meyer developed the log-polynomial adaptive-regression model to account for their results.

@@

*Model Fits to Experiment 2 and 3.*

Data from Experiments 2 and 3 were fit to ALM and EXAM in the same manner as Experiment1 . For brevity, we only plot and discuss the results of the "fit to training and testing data" models - results from the other fitting methods can be found in the appendix. The model fitting results for Experiments 2 and 3 closely mirrored those observed in Experiment 1. The Bayesian regression models predicting model error as a function of Model (ALM vs. EXAM), Condition (Constant vs. Varied), and their interaction (see @tbl-htw-ee-e23) revealed a consistent main effect of Model across all three experiments. The negative coefficients for the ModelEXAM term (Exp 2: $\beta$ = `r bm1_e2$estimate`, 95% CrI `r bm1_e2$conf.low`, `r bm1_e2$conf.high`, pd = `r bm1_e2$pd`; Exp 3: $\beta$ = `r bm1_e3$estimate`, 95% CrI `r bm1_e3$conf.low`, `r bm1_e3$conf.high`, pd = `r bm1_e3$pd`) indicate that EXAM outperformed ALM in both experiments. Furthermore, the interaction between Model and Condition was significant in both Experiment 2 ($\beta$ = `r bm3_e2$estimate`, 95% CrI `r bm3_e2$conf.low`, `r bm3_e2$conf.high`, pd = `r bm3_e2$pd`) and Experiment 3 ($\beta$ = `r bm3_e3$estimate`, 95% CrI `r bm3_e3$conf.low`, `r bm3_e3$conf.high`, pd = `r bm3_e3$pd`), suggesting that the superiority of EXAM over ALM was more pronounced for the Constant group compared to the Varied group, as was the case in Experiment 1. Recall that Experiment 3 included participants in both the original and reverse order conditions - and that this manipulation interacted with the effect of training condition. We thus also controlled for band order in our Bayesian Regression assessing the relative performance of EXAM and ALM in Experiment 3. There was a significant three way interaction between Model, Training Condition, and Band Order ($\beta$ = `r bm4_e3$estimate`, 95% CrI `r bm4_e3$conf.low`, `r bm4_e3$conf.high`, pd = `r bm4_e3$pd`), indicating that the relative advantage of EXAM over ALM was only more pronounced in the original order condition, and not the reverse order condition (see @fig-e2_e3_ae).

@@

*Computational Model Summary*.

Across all three experiments, the model fits consistently favored the Extrapolation-Association Model (EXAM) over the Associative Learning Model (ALM). This preference for EXAM was particularly pronounced for participants in the constant training conditions (note the positive coefficients on ModelEXAM:conditVaried interaction terms @tbl-htw-ee-e23). This pattern is clearly illustrated in @fig-htw-best-model, which plots the difference in model errors between ALM and EXAM for each individual participant. Both varied and constant conditions have a greater proportion of subjects better fit by EXAM (positive error differences), with the magnitude of EXAM's advantage visibly larger for the constant group.
